A Constraint Solving Approach to Parikh Images of Regular Languages

### 2022-07-07 main.tex
Work on implementation, shortening

Please note that \Catra{} uses symbolic labels for automata. A symbolic label is
defined as a range of matching single Unicode code points. This allows succinct
representation of many regular expression patterns such as \lstinline{(a-z).*}
which would have otherwise required $27$~nearly identical transitions. In this
instance, the transition would be defined as \lstinline{a -> b [97, 122]}, as the
interval is closed from both edges.

In satisfaction mode, supported by all back-ends, \Catra{} tries to satisfy the
constraints expressed by the input file, reporting satisfiable with register
assignments or unsatisfiable much like traditional~SAT- or SMT solvers would.
Additionally, the baseline and our own back-end also support generation of the
Presburger formula describing the constraints of the input file.

\Catra{} can solve equations of Parikh images. An equation like
$\ParikhMap(\Automaton_1 \times \Automaton_2) = \ParikhMap(\Automaton_3 \times
\Automaton_4)$ would be expressed as one \lstinline{synchronised} block per side of
the equation containing their respective terms followed by a
\lstinline{constraint} requiring all their counters to agree (e.g
\lstinline|Aa = Ba && Ab = Bb ...|). Note that the product construction will implicitly require the counters of each term to agree with each other (and the product), so equalities between all counters of all automata are not necessary.

We implement \Calculus{} as a theory plug-in for the \Princess{} theorem prover.
Since \Princess{} does not support multiple-arity predicates like the ones we
use in \Calculus{}, we have implemented variable-length arguments using
additional helper predicates. These are $\Unused{}(\Automaton)$, which marks an
automaton as unused in any product, and $\TransitionMask{}(\Automaton,
\Transition, \Filter(\Transition))$ which associates a transition $\Transition$
and automaton $\Automaton$ with its corresponding transition variable.
Additionally, we associate each of our predicates with an instance variable in
order to differentiate instances of the predicates.

@@

We implement the baseline approach using the same Presburger solver
(\Princess{}), input file parser, and automaton implementation as \Catra. We
do this in order to make the comparison between methods as fair as possible, and
give us the ability to compare the efficiency of the calculus rules themselves.
Using the formula of Equation~\ref{eq:generate-parikh}, we produce quantified
Presburger formulae for each successive term and add them to Princess. In this
fashion we compute the product incrementally term by term, checking
satisfiability in each step. We use a priority queue to select which
term to use for the next step of the product, and order it by the number
of transitions as a heuristic for the size of the automaton. We use this heuristic
to avoid computing large (and therefore slow) products until we have to, banking
our hopes on finding a source of unsatisfiability early. The pseudocode for
our implementation can be seen in Algorithm~\ref{alg:baseline}.

@@

It should be noted that our automata (including the successive products) are
always by construction forward- and backward- reachable-minimal. We avoid
computing fully minimal automata since it is unclear when an automaton with
registers can be safely minimised. This guarantees that any automaton we produce
only contains states that are both reachable from the initial state and has a
path to an accepting state. We never perform any other minimisation on the
automata for either backend. More complex minimisation was left out since
performing minimisation on automata with counters is non-trivial, and indeed
already minimising symbolic automata is itself risks exponential blowup
\cite{minimising-symbolic}.

@@

There are a number of choices left unspecified in \Calculus{} as described in
Sections~\ref{sec:calculus} and~\ref{sec:multiple}. For example, the order of
materialisation of intermediate products and the order of splitting. In
addition, the splitting specified in the general calculus is correct but
inefficient. In this section we describe additional implementation details used
to enhance our solution.

@@

We order our rule applications so that we first propagate connectedness if
possible, then perform materialisation if tractable (see below!), then finally
resort to splitting if we must.

In addition to the splitting on individual transition variables described in Table~\ref{tbl:rules:single}, where we randomly select a variable whose bounds are not known, we also prefer splitting to sever/include a strongly connected component. We randomly select an automaton where we can compute a cut to at least one strongly connecting component that can be severed from the initial state, e.g. where the strongly connected component does not contain the initial state and where the sum of the transition variables we have associated with the transitions of a minimal cut between the initial state and an arbitrary state in the component is not known to be greater than 0, but also not equal to 0. If there are multiple such strongly connected components we choose one randomly. We then proceed to split on the cut as if it were a regular transition, e.g. its sum being zero or nonzero. In this way we use our splitting strategy to drive the connectedness constraint towards propagation by cutting off loops from the automata.

@@

Product materialisation, then, is the final piece of the puzzle. In the current implementation, we put off computing intermediate products until at least all but 2 transition variables of one of them have a known status, e.g. either is known to be zero or known to be positive. The number was chosen experimentally, and we observe a consistent trend that to have a low number and therefore a high threshold for materialisation is better throughout the implementation of other features. The other automaton for the product is selected randomly.

@@

\Catra{} enables clause learning by default when using our backend, as it has
been experimentally shown to increase the performance in aggregate (though not
strictly). We do not currently implement all the proposed features of
\cref{sec:ext:backjumping}, but we do implement forward-reachability cut
learning (which had a very modest improvement in performance). No sophisticated
clause learning for products has been implemented.

@@

Finally, we perform restarts scaled by the Luby series~\cite{luby}. Experimental
results have shown this to have a large improvement in performance, which is
unsurprising given how many random choices we make during solving and how
tail-heavy our problem is. It seems to be very easy to get stuck with a poor
choice of product to materialise or transitions to split on.

@@

In \cref{fig:solve-division} we show how many of the~\NrBenchmarks{} instances
the respective back-ends could solve and with which status. A full summary of
their outcomes is also available in \cref{tab:solve-status}. We see that
\Calculus{} generally outperforms \Nuxmv{} on unsatisfiable instances, while
being worse at satisfiable instances. Both \Nuxmv{} and \Calculus{} strictly
outperform baseline on every kind of instance, but most of all on satisfiable
instances. \Cref{tab:solve-status} also shows a small number of bugs in
\Calculus{}: 27~satisfiable instances are misreported as \Unsat{}, and 20
additional instances fail with an error. Both problems are due to
non-deterministic bugs in the clause learning. Note that \Calculus{} is sound:
it never misreports an unsatisfiable instance as \Sat{}, and all
\Fudge{\numprint{6218}} satisfying assignments have been verified by \Nuxmv{}.

>>>>

\Catra{} uses symbolic labels for automata. A symbolic label is defined as a
finite range of Unicode code points. This allows succinct representation of many
regular expression patterns such as \lstinline{(a-z).*} which would have
otherwise required $27$~transitions. The transition for the range would be
written \lstinline{a -> b [97, 122]}.

In satisfaction mode, supported by all backends, \Catra{} tries to satisfy the
constraints expressed by the input file, reporting \Sat{} with register
assignments or \Unsat{} much like traditional~SAT- or SMT solvers would.
Additionally, baseline and \Calculus{} also support generation of the Presburger
formula describing the constraints of the input file. Baseline uses standard
quantifier elimination, and \Calculus{} uses the method described in
\cref{sec:finding-the-image}.

Since \Princess{} does not support multiple-arity predicates like the ones we
use in \Calculus{}, we have implemented variable-length arguments using
additional helper predicates. These are $\Unused{}(\Automaton)$, which marks an
automaton as unused in any product, and $\TransitionMask{}(\Automaton,
\Transition, \Filter(\Transition))$ which associates a transition $\Transition$
and automaton $\Automaton$ with its corresponding transition variable.
Additionally, we associate each of our predicates with an instance variable in
order to differentiate instances of the predicates.

@@

We baseline using the same Presburger solver (\Princess{}), input file parser,
and automaton implementation as \Catra. We do this in order to better analyse
the impact of the calculus rules themselves. Using the formula of
\eqref{eq:generate-parikh}, we produce quantified Presburger formulae for each
successive term and add them to \Princess. We compute the product incrementally
term by term, checking satisfiability at each step. We use a priority queue to
select automata for each step, and order it by the number of transitions as a
heuristic for the size of the automata. We use this heuristic to avoid computing
large (and therefore slow) products until we have to, banking our hopes on
computing an empty intermittent product early. The pseudocode for our
implementation can be seen in \cref{alg:baseline}.

@@

As an optimisation, our automata (including intermittent products) are created
forward- and backward- reachable-minimal. Any automaton we produce only contains
states that are both reachable from the initial state and has a path to an
accepting state. We never perform any other minimisation on the automata for
either backend. More complex minimisation was left out since performing
minimisation on automata with counters is non-trivial, and minimising symbolic
automata risks exponential blowup \cite{minimising-symbolic}.

@@

There are a number of choices left unspecified in \Calculus{} as described in
\cref{sec:calculus,sec:multiple}. For example, the order of materialisation of
intermediate products and the order of splitting. In this section we describe
additional implementation details and techniques used to enhance \Catra.

@@

We order our rule applications so that we first propagate connectedness if
possible, then perform materialisation if tractable as defined below, then
finally resort to splitting if we must.

In addition to applying \Split{} as described in \cref{tbl:rules:single} to
randomly selected transitions, we prefer splitting to sever a strongly connected
from the initial state. We randomly select an automaton where we can compute a
cut between an SCC and the initial state, that is where the SCC does not contain
the initial state and where the sum of the transition variables of the
transitions in the cut is not known to be positive. If there are multiple such
strongly connected components we choose one randomly. We then proceed to split
on the sum of the transition variables of the cut as if it were a regular
transition, e.g. its sum being zero or nonzero. In this way we drive \Calculus{}
towards applying \Propagate{}.

@@

Product materialisation is the final piece of the puzzle. In the current
implementation we put off computing intermediate products until at least all but
two transition variables of one of the automata is known to be either present or
absent. The number was chosen experimentally, and we observe a consistent trend
towards lower numbers being better. The other automaton for the product is
selected randomly.

@@

\Catra{} enables clause learning by default when using our backend, as it has
been experimentally shown to increase the performance in aggregate (though not
strictly). We do not currently implement all the proposed features of
\cref{sec:ext:backjumping}, but we do implement forward-reachability cut
learning. No sophisticated clause learning for products has been implemented.

@@

Finally, we perform restarts scaled by the Luby series~\cite{luby}. Experimental
results have shown this to have a large improvement in performance, which is
unsurprising given how many random choices we make during solving and how
tail-heavy our problem is.

@@

In \cref{fig:solve-division} we show how many of the~\NrBenchmarks{} instances
the respective back-ends could solve and with which status. A full summary of
their outcomes is also available in \cref{tab:solve-status}. We see that
\Calculus{} generally outperforms \Nuxmv{} on unsatisfiable instances, while
being worse at satisfiable instances. Both \Nuxmv{} and \Calculus{} strictly
outperform baseline on every kind of instance, but most of all on satisfiable
instances. \Cref{tab:solve-status} also shows a small number of bugs in
\Calculus{}: 27~satisfiable instances are misreported as \Unsat{}, and 20
additional instances fail with an error. Both problems are due to
non-deterministic bugs in the clause learning. Note that \Calculus{} is sound:
it never misreports an unsatisfiable instance as \Sat{}, and all
\Fudge{\numprint{6218}} satisfying assignments have been verified by \Nuxmv{}.
Baseline performs worse on satisfiable instances because it executes a heuristic
meant to detect unsatisfiability early at a heavy penalty to satisfiable
instances.
